Working Principles of Proof Assistants and Formalization of some proofs in Agda

### 2025-08-05 Mainpaper/paper.tex
REPORT DONE

% ======== Packages =========
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{geometry}
\usepackage{bussproofs}
\usepackage{braket}
\geometry{a4paper, margin=1in}
\usepackage{titlesec}
\usepackage{newunicodechar}
\usepackage{tikz}
\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}    % ∀
\DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}       % ℓ
\DeclareUnicodeCharacter{2192}{\ensuremath{\rightarrow}}% →
\DeclareUnicodeCharacter{00D7}{\ensuremath{\times}}     % ×
\DeclareUnicodeCharacter{228E}{\ensuremath{\uplus}}   % ⊎
\DeclareUnicodeCharacter{22A5}{\ensuremath{\bot}}       % ⊥
\DeclareUnicodeCharacter{00AC}{\ensuremath{\neg}}       % ¬
\DeclareUnicodeCharacter{2081}{\textsubscript{1}}       % ₁
\DeclareUnicodeCharacter{2082}{\textsubscript{2}}       % ₂
\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}    % λ
\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}   % Σ
\geometry{
  left=1.2in,
  right=1.2in,
  top=1.1in,
  bottom=1.1in
}
\usetikzlibrary{arrows.meta, positioning}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

@@

        \includegraphics[width=4cm]{logo.png}\\[2cm]
        % Replace with your logo file. Place 'ku_logo.png' in the same folder!

@@

In 1998 Thomas C. Hales announced that he had proved the kepler conjecture
(Solane,1998) However the peer review took over 4 years especially due to the proof
being incredibely difficult to check, with over a dozen of mathematician to refree
the proof and although they were 99% certain it was not untill the proof became
formalized that they were able to completely validate the correctness of the proof,
which according to Thomas Hales, would have taken 20 person-year of manual work.
[Szpiro, G, 2003]

Proof assistants helps with formalization of mathematical proofs in computer which
enables for their verification digitally. Some exhaustive proofs may contain large
number of cases which is not feasible to write and verify manually and requires
computer assitance.
Such Proof assistants (systems) work by treating proofs as computational objects
and checking them using strict logical rules. To enable this, proofs must be writ-
ten in a formal language that removes ambiguity and allows machines to interpret
them accurately. Underlying this process is a theoretical framework—often type
theory—that defines how propositions and proofs relate to each other.
This method bridges programming and logic: writing a proof resembles writing a
program, and verifying it is like type-checking. By interpreting logic computation-
ally, proof assistants ensure that every proof is exact, complete, and verifiable by
a machine. This project invloves investigating this process.

\subsection{Objectives}
\begin{enumerate}
    \item To explore the correspondence between Proofs and Programs, Type Theory and Intuitionstic Logic.
    \item To investigate a current exsisting proof assistants (Agda) and identify the mathematics behind it.
    \item To formalize some selected proofs in Agda and demonstrate verification process.
\end{enumerate}

@@

\subsection{Defining Natural Numbers, Equality and Proof of Assocaitive and Commutative Properties}
\begin{verbatim}

@@

\end{verbatim}
\subsection{Euclid's Theorem of Primes }

\subsection{Euclid's Algorithm's Proof}

\subsection{Attempt to Constructive Reals}

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{geometry}
\usepackage{bussproofs}
\usepackage{braket}
\geometry{a4paper, margin=1in}
\usepackage{titlesec}
\usepackage{newunicodechar}
\usepackage{tikz}
\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}    % ∀
\DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}       % ℓ
\DeclareUnicodeCharacter{2192}{\ensuremath{\rightarrow}}% →
\DeclareUnicodeCharacter{00D7}{\ensuremath{\times}}     % ×
\DeclareUnicodeCharacter{228E}{\ensuremath{\uplus}}   % ⊎
\DeclareUnicodeCharacter{22A5}{\ensuremath{\bot}}       % ⊥
\DeclareUnicodeCharacter{00AC}{\ensuremath{\neg}}       % ¬
\DeclareUnicodeCharacter{2081}{\textsubscript{1}}       % ₁
\DeclareUnicodeCharacter{2082}{\textsubscript{2}}       % ₂
\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}    % λ
\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}   % Σ
\DeclareUnicodeCharacter{211A}{\ensuremath{\mathbb{Q}}} % ℚ
\DeclareUnicodeCharacter{2223}{\ensuremath{\mid}}       % ∣
\DeclareUnicodeCharacter{2264}{\ensuremath{\leq}}       % ≤
\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}     % ≡
\DeclareUnicodeCharacter{21D2}{\ensuremath{\Rightarrow}}% ⇒
\DeclareUnicodeCharacter{2243}{\ensuremath{\simeq}}     % ≃
\DeclareUnicodeCharacter{2124}{\ensuremath{\mathbb{Z}}} % ℤ
\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}} % ℕ
\DeclareUnicodeCharacter{211D}{\ensuremath{\mathbb{R}}} % ℝ

\geometry{
  left=1.2in,
  right=1.2in,
  top=1.1in,
  bottom=1.1in
}
\usetikzlibrary{arrows.meta, positioning}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

@@

        \includegraphics[width=4cm]{logo.png}\\[2cm]

@@

In 1998 Thomas C. Hales announced that he had proved the kepler conjecture
(Solane,1998) However the peer review took over 4 years especially due to the proof
being incredibly difficult to check, with over a dozen of mathematician to referee 
the proof and although they were 99 percent certain it was not until the proof became
formalized that they were able to completely validate the correctness of the proof,
which according to Thomas Hales, would have taken 20 person-year of manual work.
[Szpiro, G, 2003]

Proof assistants helps with formalization of mathematical proofs in computer which
enables for their verification digitally. Some exhaustive proofs may contain large
number of cases which is not feasible to write and verify manually and requires
computer assistance.
Such Proof assistants (systems) work by treating proofs as computational objects
and checking them using strict logical rules. To enable this, proofs must be writ-
ten in a formal language that removes ambiguity and allows machines to interpret
them accurately. Underlying this process is a theoretical framework—often type
theory—that defines how propositions and proofs relate to each other.
This method bridges programming and logic: writing a proof resembles writing a
program, and verifying it is like type-checking. By interpreting logic computation-
ally, proof assistants ensure that every proof is exact, complete, and verifiable by
a machine. This project involves investigating this process.

\subsection{Objectives}
\begin{enumerate}
    \item To explore the correspondence between Proofs and Programs, Type Theory and Intuitionstic Logic.
    \item To investigate a current existing proof assistants (Agda) and identify the mathematics behind it.
    \item To formalize some selected proofs in Agda and demonstrate verification process.
\end{enumerate}

@@

\subsection{Defining Natural Numbers, Equality and Proof of Associative and Commutative Properties}
\begin{verbatim}

@@

\subsection{Attempt to Formalize Constructive Reals}
\begin{verbatim}
    module rls where
open import Relation.Binary.Core using (Rel)

open import Data.Rational as ℚ using (ℚ; -_ ; _*_; 
  _-_; _+_; ∣_∣; _≤_; *≤* ; _≤?_; _÷_; ≡⇒≃)

open import Data.Integer as ℤ using (ℤ; +_ ; -[1+_]; pos; _%_)
open import Data.Nat as ℕ using (ℕ; suc; zero; _≤?_)
import Level

natToInteger : ℕ → ℤ
natToInteger zero    = 0
natToInteger (suc n) = pos (pos (suc n))

oneℤ : ℤ
oneℤ = pos 1

reciprocal : ℕ → ℚ
reciprocal zero    = 0 % 1
reciprocal (suc n) = oneℤ % pos (suc n)

record ℝ : Set where
  constructor Real
  field
    f : ℕ -> ℚ
    reg : {n m : ℕ} -> ∣ f n - f m ∣ ≤ reciprocal(suc n) + reciprocal(suc m)

_≃_ : Rel ℝ Level.zero
x ≃ y =  {n : ℕ} -> ∣ ℝ.f x n - ℝ.f y n ∣ ≤ reciprocal(suc n) + reciprocal(suc n)

@@

\begin{itemize}
    \item In 5.1 we did everything from scratch, starting from definitons by ourselves. For larger and more proofs doing this for every .agda code is difficult. We make use of Agda Standard Library and leverage some pre defined objects and pre proven theorems.
    \item In 5.3, For the second part of the therorem, we had take non constructive approach, which was done by using Dec from Relation.Nullary, which is the law of excluded middle.
    \item In 5.4, We tried to construct real numbers as Cauchy sequences of Rational Numbers like in \cite{BB85}, where we encountered a problem while creating a reciprocal function $\mathbb{N} \to \mathbb{Q}$. The issue being in Agda the numerator requires to be a integer, and we did not find a way to introduce the integer 0. This remains unresolved.
\end{itemize}
